---
author: Beier Jiang
title: Comparative Evaluation of Parametric and Nonparametric Models
subtitle: for Health and Functional Limitations Prediction
institute: McMaster University
date: last-modified
date-format: "[Updated] MMMM DD, YYYY"
editor: source
## Pass knitr options like this (otherwise old-school knitr::opt_chunks$set() in an R chunk)
## Squirrly issues, trying setting autodep and cache in opts_chunk below
knitr:
  opts_chunk:
    autodep: false
    cache: true
    collapse: true
    dev: tikz
    echo: false
    message: false
    warning: false
    R.options:
      np.messages: false
      crs.messages: false
      plot.par.mfrow: false
## revealjs options galore
format:
  revealjs:
    background-transition: fade
    center: true
    chalkboard: true
    code-link: true
    css: custom.css
    embed-resources: false
    footer: "ECON 768 | &copy; Beier Jiang"
    ## This "hack" is exclusive to using mathjax and allows for plain vanilla
    ## equation cross referencing using any LaTeX math environment such as
    ## align, equation, split, etc., i.e., \label{eqfoo} and \eqref{eqfoo} (it
    ## can also be used for revealjs, html, while standard pdf will naturally
    ## accept it using defaults - breaks if used for docx, ppt etc.)
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          window.MathJax = {
            tex: {
              tags: 'ams'
            }
          };
          </script>
    incremental: true
    link-external-newwindow: true
    ## multiplex: true will create two html files:
    ## 1. index.html: This is the file you should publish online and that your
    ## audience should view.
    ## 2. index-speaker.html: This is the file that you should present from.
    ## This file can remain on your computer and does not need to be published
    ## elsewhere.
    multiplex: true
    preview-links: auto
    self-contained-math: false
    show-notes: false
    slide-number: true
    theme: default
    touch: true
    transition: slide
---

# Slide Pro-Tips

::: nonincremental
-   View **full screen** in A2L by clicking on the slide and then pressing the F key (press the Esc key to revert)

-   Access **navigation menu** by pressing the M key (click X in navigation menu to close)

-   **Advance** using arrow keys

-   **Zoom** in by holding down the Alt key in MS Windows, Opt key in macOS or Ctrl key in Linux, and clicking on any screen element (Alt/Opt/Ctrl click again to zoom out)

-   Use **copy to clipboard** button for R code blocks (upper right in block) to copy and paste into R/RStudio

-   **Export to a PDF** by pressing the E key (wait a few seconds, then print \[or print using system dialog\], enable landscape layout, then save as PDF - press the E key to revert)

-   Enable drawing tools - chalk **board** by pressing the B key (B to revert), notes **canvas** by pressing the C key (C to revert), press the Del key to erase, press the D key to **download drawings**
:::

```{r global_options, include=FALSE}
require(pdftools)
pdf2png <- function(path) {
  # only do the conversion for non-LaTeX output
  if (knitr::is_latex_output())
    return(path)
  path2 <- xfun::with_ext(path, "png")
  img <- magick::image_read_pdf(path)
  magick::image_write(img, path2, format = "png")
  path2
}
## This function takes a confusion matrix and formats it correctly if
## it is unbalanced and returns the CCR as well.
CM <- function(cm) {
  factor.values.eval <- colnames(cm)
  CM <- matrix(0,nrow(cm),nrow(cm))
  rownames(CM) <- rownames(cm)
  colnames(CM) <- rownames(cm)
  for(i in 1:ncol(cm)) CM[,(1:nrow(cm))[rownames(cm)==factor.values.eval[i]]] <- cm[,i]
  return(list(CM=CM,CCR=sum(diag(CM))/sum(CM)))
}
```

```{r pdf2png_options, include=FALSE, cache=FALSE}
## Set some options for knitr to apply globally
knitr::opts_chunk$set(fig.process=pdf2png)
## Don't cache libraries or data
library(np)
require(sem)
library(crs)
library(MASS)
```

## Data

-   The data is from wave 10 (2021/23) of the English Longitudinal Study of Ageing (ELSA)
-   Compare the classification ability of parametric Probit model, Probit interaction model, BIC-optimal model and nonparametric model
-   In ELSA, the questionnaire asks the respondents about their self-reported health (SRH) and difficulties with activities of daily living (ADLs) as well as instrumental ADLs (IADLs)
-   This application has n = 7117 and 6 explanatory variables
    -   *health* indicator of poor health on SRH (1 = fair or poor)
    -   *ADL* indicator of difficulties in at least one of 5 ADLs (1 = yes)
    -   *IADL* indicator of difficulties in at least one of 5 IADLs (1 = yes)

## Data

-   Definitions of 6 variables:
    -   *sex* gender of the respondent (1 = male)
    -   *work* current working status (1 = working)
    -   *couple* current marital status (1 = married)
    -   *ed* educational attainment (1 = still in school, 2 = over 18, 3 = 18 and under, 4 = no school)
    -   *hh* number of adults and children in a household
    -   *age* the age of people 50 and older in years
-   Note that all variables other than age and hh are categorical in nature

## Results - Health

```{r table_health, echo=FALSE}
library(knitr)
data_health <- data.frame(
  Model = c("Linear", "Lin-Int", "BIC", "Kernel"),
  Apparent = c(0.727, 0.734, 0.728, 0.735),
  Expected = c(0.724, 0.729, 0.728, 0.731),
  Rank = c(4, 2, 3, 1)
)
kable(data_health, caption = "Table 1: Health Model Performance", format = "html") 

```

-   The results only use n = 2000
-   The kernel model has the highest expected true CCR, so it performs best

## Results - ADL

```{r table_adl, echo=FALSE}
data_adl <- data.frame(
  Model = c("Linear", "Lin-Int", "BIC", "Kernel"),
  Apparent = c(0.836, 0.836, 0.836, 0.840),
  Expected = c(0.837, 0.833, 0.838, 0.837),
  Rank = c(2, 4, 1, 3)
)
kable(data_adl, caption = "Table 2: ADL Model Performance", format = "html")

```

-   The BIC model has the highest expected true CCR
-   This may be because the true relationship between the predictor and ADL is linear

## Results - IADL

```{r table_iadl, echo=FALSE}
data_iadl <- data.frame(
  Model = c("Linear", "Lin-Int", "BIC", "Kernel"),
  Apparent_CCR = c(0.904, 0.904, 0.904, 0.905),
  Expected_CCR = c(0.905, 0.904, 0.905, 0.904),
  Rank = c("1.5", "3.0", "1.5", "4.0")
)
kable(data_iadl, caption = "Table 3: IADL Model Performance", format = "html")

```

-   The linear and BIC models have the highest expected true CCR, but the difference between the other models is only 0.1%
-   This shows that increased flexibility (e.g., interaction or higher-order terms) does not significantly improve performance compared to a simple linear model

## Conclusion

-   The results show that the nonparametric model has a good classification of *health* only
-   This could be due to the fact that the study consolidated the data, which reduced variation in the data
-   As a result, only BIC-optimal model stand out in all three dependent variables

## References

Fonseca, R., Michaud, P. C., and Zheng, Y. (2020). The effect of education
on health: evidence from national compulsory schooling reforms. SERIEs,
11:83–103.